---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

### load data 
```{r loading} 
data <- read.csv(unzip("activity.zip", "activity.csv"))
dim(data)
```

## What is mean total number of steps taken per day?

#### Obtain total of steps for each day
```{r histogram }
s <- split(data, data$date)
s.sum<-sapply(s, function(r) { sum(r$steps,na.rm=T)})
hist(s.sum, main="Total steps for each day", xlab="Total steps")
```

#### Mean and median of total steps per day

```{r }
mean(s.sum)
median(s.sum)
```

## What is the average daily activity pattern?

Generate a plot of a time series containing the mean number of steps for each interval across all days 
```{r}
ts <- sapply(split(data$steps, data$interval), function(x) { mean(x, na.rm=T) } )
plot(attributes(ts)[[1]], ts, type="l", main="Mean number of steps / intervall across all days", 
    xlab="Intervals", ylab="Avg #steps")
```

Find the interval in the time series above with maximum average number of steps 

```{r}
attributes(ts)[[1]][which.max(ts)]
```

## Inputing missing values

The number of good values and the number of missing values are
```{r}
r.good <- sum(!is.na(data$steps))
r.missing <- sum(is.na(data$steps))
```

The number of good readings is `r r.good` and of missing values is `r r.missing`.
The percentage of missing values is 
```{r}
r.missing/r.good*100
```

Checking the distribution of missing (1) values against good values (2)

```{r}
sapply(s, function (x) { c(sum(is.na(x$steps)), sum(!is.na(x$steps))) })
```

We can see that missing values involve all readings of a day. 

Therefore we have to fill all intervals of such days with estimated values. 
The best approach is fill the gap using the mean number of steps/intervals for the corresponding day of the week.

#### Steps to estimate values for missing days

1. Include weekdays in main datasource
```{r}
data$weekday <- weekdays(strptime(data[,2],"%Y-%m-%d"))
```

2. Split datasets for each weekday
```{r}
s.wday <- split(data, data$weekday)
```

3. Obtain names of weedays
```{r}
w.names <- attributes(s.wday)[[1]]
```

4. Compute time series of mean number of steps / interval for corresponding weekday 
```{r}
ts.w <- list()
i <- 0 
for (v in s.wday) { 
    i <- i+1;
    ts.w[[i]] <- sapply(split(v$steps, v$interval), function(r) { mean (r, na.rm=T)} )  
}
```

5. The names of weekdays obtanined in (3) correspond to the time series obtained in (4). Define a function to map dates -> weekdays (3) -> index of the time series calculated in 4
```{r}
# x: string containing date in the format YYYY-mm-dd
# w.names: vector containing days of week, whose indexes correspond to tables with 
# average number of steps for that day of the week / interval

get.day.index <- function (x, w.names) {
    x.wd <- weekdays(strptime(x, "%Y-%m-%d")); 
    which(w.names==x.wd)
}
```

6. Finally, create a replica of the original data filling in values for days with missing readings using estimated values for the corresponding weekday
```{r}
s.new <- s
for (i in 1:length(s.new)) {
    is.missing <- sum(is.na(s.new[[i]]$steps)) != 0    
    if (is.missing) {
        x <- s.new[[i]]
        day.index <- get.day.index(x$date[1],w.names)
        x$steps <- ts.w[[day.index]]
        s.new[[i]] <- x
    }
}
data.new <- unsplit(s.new, data$date)
```

7. Plot the histogram with the total of steps per each day using dataset containing estimated values

```{r}
s.new.sum<-sapply(s.new, function(r) { sum(r$steps,na.rm=T)})
hist(s.new.sum, main="Total steps for each day (with estimations for missing values)", xlab="Total steps")
```

and calculate mean and median over the new dataset.

```{r}
mean(s.new.sum)
median(s.new.sum)
```

## Are there differences in activity patterns between weekdays and weekends?

First, set a weekend flag in the main datasource
```{r}
data$isWeekend <- data$weekday %in% c("Saturday", "Sunday")
```

And calculate a time series for weekends and weekdays

```{r}
ts.weekend <- with(data[data$isWeekend,], sapply(split(steps, interval), function(x) { mean(x, na.rm=T) }))
ts.weekday <- with(data[!data$isWeekend,], sapply(split(steps, interval), function(x) { mean(x, na.rm=T) }))
```

Show the plot for both time series
```{r}
par(mfrow = c(1, 2))
plot(attributes(ts.weekday)[[1]], ts.weekday, type="l", main="Weekdays", xlab="Intervals", ylab="Average amount of steps")
plot(attributes(ts.weekend)[[1]], ts.weekend, type="l", main="Weekends", xlab="Intervals", ylab="Average amount of steps")
```
